Formalization of Typed Lua

### 2015-06-11 doc/papers/paper/paper.tex
rewrite of the section on subtyping for DLS paper

The rest of the paper is organized as follows:
Section \ref{sec:statistics} presents the methodology and results of
our survey of Lua idioms;
Section \ref{sec:types} presents the types of Typed Lua's
type system, and their subtyping and consistent-subtyping relations;
Section \ref{sec:rules} presents the typing rules for evolution of table types,
and the typing rules for destructuring unions of tuples into projection types,
allowing flow typing on these projections;
Section \ref{sec:related} reviews related work;
Section \ref{sec:conclusion} presents our conclusions,
and outlines our plans for future work.

@@

\section{Types and Subtyping}
\label{sec:types}

@@

Second-level types are either tuple types or unions of tuple types, and represent sequences of values that appear as arguments, return values, and the left and right sides of
assignments. Tuple types are heterogeneous lists of first-class
types that end either with $\Void$ or with a {\em variadic type}.
A variadic type $T{*}$ represents an unbounded sequence of values of type $T \cup \Nil$.

@@

\subsection{Subtyping}

Our type system uses subtyping \cite{cardelli1984smi,abadi1996to} to order
types and consistent-subtyping \cite{siek2007objects,siek2013mutable}
to allow the interaction between statically and dynamically typed code.
We explain the subtyping and consistent-subtyping rules throughout this section.
However, we focus the discussion on the definition of subtyping because
we can combine the consistency and subtyping relations to achieve
consistent-subtyping \cite{siek2007objects,siek2013mutable}.
The differences between consistent-subtyping and subtyping are the way
they handle the dynamic type, and the fact that subtyping is transitive,
but consistent-subtyping is not.

We present the subtyping rules as a deduction system for the
subtyping relation $\senv \vdash T_{1} \subtype T_{2}$.
The variable $\senv$ is a set of pairs of recursion variables.
We need this set to record the hypotheses that we assume when checking
recursive types.

The subtyping rules for literal types and base types include the rules
for defining that literal types are subtypes of their respective base types,
and that $\Integer$ is a subtype of $\Number$:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-FALSE}\\
\senv \vdash \False \subtype \Boolean
\end{array}
\;
\begin{array}{c}
\mylabel{S-TRUE}\\
\senv \vdash \True \subtype \Boolean
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-STRING}\\
\senv \vdash {\it string} \subtype \String
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-INT1}\\
\senv \vdash {\it int} \subtype \Integer
\end{array}
\;
\begin{array}{c}
\mylabel{S-INT2}\\
\senv \vdash {\it int} \subtype \Number
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FLOAT}\\
\senv \vdash {\it float} \subtype \Number
\end{array}
\;
\begin{array}{c}
\mylabel{S-INTEGER}\\
\senv \vdash \Integer \subtype \Number
\end{array}
\end{array}
\]

Subtyping is reflexive and transitive;
therefore, we could have omitted the rule \textsc{S-INT2}.
More precisely, we could have defined a transitive rule for first-level
types instead of defining specific rules for transitive cases.
For instance, a transitive rule would allow us to derive that
\[
\dfrac{\senv \vdash 1 \subtype \Integer \;\;\;
       \senv \vdash \Integer \subtype \Number}
      {\senv \vdash 1 \subtype \Number}
\]

However, we are using the subtyping rules as the template for defining
the consistent-subtyping rules, and consistent-subtyping is not
transitive.
More precisely, we want the subtyping and consistent-subtyping rules
to differ only in the way they handle the dynamic type.
Thus, we define the subtyping rules using an algorithmic approach
that is close to the implementation, as this approach allows us to use
subtyping to easily formalize consistent-subtyping.

Our type system includes the top type $\Value$,
so any first-level type is a subtype of $\Value$:
\[
\begin{array}{c}
\mylabel{S-VALUE}\\
\senv \vdash T \subtype \Value
\end{array}
\]

Another type that is only a subtype of itself and of the type $\Value$
is the type $\Self$.

The subtyping rules for union types are standard:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-UNION1}\\
\dfrac{\senv \vdash T_{1} \subtype T \;\;\;
       \senv \vdash T_{2} \subtype T}
      {\senv \vdash T_{1} \cup T_{2} \subtype T}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-UNION2}\\
\dfrac{\senv \vdash T \subtype T_{1}}
      {\senv \vdash T \subtype T_{1} \cup T_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-UNION3}\\
\dfrac{\senv \vdash T \subtype T_{2}}
      {\senv \vdash T \subtype T_{1} \cup T_{2}}
\end{array}
\end{array}
\]

The first rule shows that a union type $T_{1} \cup T_{2}$
is a subtype of $T$ if both $T_{1}$ and $T_{2}$ are subtypes
of $T$;
and the other rules show that a type $T$ is a subtype
of a union type $T_{1} \cup T_{2}$ if $T$ is a subtype of
either $T_{1}$ or $T_{2}$.

The subtyping rule for function types is also standard:
\[
\begin{array}{c}
\mylabel{S-FUNCTION}\\
\dfrac{\senv \vdash S_{3} \subtype S_{1} \;\;\;
       \senv \vdash S_{2} \subtype S_{4}}
      {\senv \vdash S_{1} \rightarrow S_{2} \subtype S_{3} \rightarrow S_{4}}
\end{array}
\]

The rule \textsc{S-FUNCTION} shows that subtyping between
function types is contravariant on the type of the parameter list
and covariant on the return type.
In the previous section we explained why our type system uses
second-level types to represent the type of the parameter list
and the return type.
Now, we explain their subtyping rules.

The type $\Void$ is a subtype of itself and of a variadic type:
\[
\begin{array}{c}
\mylabel{S-VOID}\\
\senv \vdash \Void \subtype T{*}
\end{array}
\]

A variadic type $T{*}$ represents a sequence of values of type
$T \cup \Nil$, and the rule \textsc{S-VOID} handles the case where
a given sequence is empty.

The subtyping rule for pair types is the standard covariant rule:
\[
\begin{array}{c}
\mylabel{S-PAIR}\\
\dfrac{\senv \vdash T_{1} \subtype T_{2} \;\;\;
       \senv \vdash P_{1} \subtype P_{2}}
      {\senv \vdash T_{1} \times P_{1} \subtype T_{2} \times P_{2}}
\end{array}
\]

The subtyping rules for variadic types are not so obvious.
We need six different subtyping rules for variadic types
to handle all the cases where they can appear.

The rule \textsc{S-VARARG1} is a special rule for handling the
case where we give a sequence of $\Nil$ to the empty tuple:
\[
\begin{array}{c}
\mylabel{S-VARARG1}\\
\senv \vdash \Nil{*} \subtype \Void
\end{array}
\]

The rule \textsc{S-VARARG2} handles the case where both tuple types end
with variadic types:
\[
\begin{array}{c}
\mylabel{S-VARARG2}\\
\dfrac{\senv \vdash T_{1} \cup \Nil \subtype T_{2} \cup \Nil}
      {\senv \vdash T_{1}{*} \subtype T_{2}{*}}
\end{array}
\]

This rule shows that $T_{1}{*}$ is a subtype of $T_{2}{*}$
if $T_{1} \cup \Nil$ is a subtype of $T_{2} \cup \Nil$.
It explicitly includes $\Nil$ in both sides because otherwise
$\Nil{*}$ would not be a subtype of several other variadic types.
For instance, $\Nil{*}$ would not be a subtype of $\Number{*}$,
as $\Nil \not\subtype \Number$.

The other rules handle the cases where only one tuple type ends with a variadic type:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-VARARG3}\\
\dfrac{\senv \vdash T_{1} \cup \Nil \subtype T_{2}}
      {\senv \vdash T_{1}{*} \subtype T_{2} \times \Void}
\end{array}
\;
\begin{array}{c}
\mylabel{S-VARARG4}\\
\dfrac{\senv \vdash T_{1} \subtype T_{2} \cup \Nil}
      {\senv \vdash T_{1} \times \Void \subtype T_{2}{*}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-VARARG5}\\
\dfrac{\senv \vdash T_{1}{*} \subtype T_{2} \times \Void \;\;\;
       \senv \vdash T_{1}{*} \subtype P_{2}}
      {\senv \vdash T_{1}{*} \subtype T_{2} \times P_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-VARARG6}\\
\dfrac{\senv \vdash T_{1} \times \Void \subtype T_{2}{*} \;\;\;
       \senv \vdash P_{1} \subtype T_{2}{*}}
      {\senv \vdash T_{1} \times P_{1} \subtype T_{2}{*}}
\end{array}
\end{array}
\]

Note that the case where both tuple types end with the type $\Void$ does
not require any special rule.
We use the subtyping rules for variadic types along with the types
$\Value$ and $\Nil$ to make our type system reflect the semantics of
Lua on discarding extra parameters and replacing missing parameters.

The subtyping rules for unions of tuple types are similar to the
subtyping rules for unions of first-level types:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-UNION4}\\
\dfrac{\senv \vdash S_{1} \subtype S \;\;\;
       \senv \vdash S_{2} \subtype S}
      {\senv \vdash S_{1} \sqcup S_{2} \subtype S}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-UNION5}\\
\dfrac{\senv \vdash S \subtype S_{1}}
      {\senv \vdash S \subtype S_{1} \sqcup S_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-UNION6}\\
\dfrac{\senv \vdash S \subtype S_{2}}
      {\senv \vdash S \subtype S_{1} \sqcup S_{2}}
\end{array}
\end{array}
\]

Back to the subtyping rules between first-level types,
the subtyping rule among a \emph{fixed} or \emph{closed}
table type and another \emph{closed} table type resembles the
standard subtyping rule between records:
\[
\begin{array}{c}
\mylabel{S-TABLE1}\\
\dfrac{\begin{array}{c}
       \forall i \in 1..n \; \exists j \in 1..m \\
       \senv \vdash K_{j} \subtype K_{i}' \;\;\;
       \senv \vdash K_{i}' \subtype K_{j} \;\;\;
       \senv \vdash V_{j} \subtype_{c} V_{i}'
       \end{array}}
      {\begin{array}{c}
       \senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{fixed|closed} \subtype\\
                    \{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{closed}
       \end{array}} \; m \ge n
\end{array}
\]

The rule \textsc{S-TABLE1} allows width subtyping and introduces the
auxiliary relation $\subtype_{c}$ to handle depth subtyping on the
type of the values stored in the table fields.
We need an auxiliary relation because the subtyping of the
type of the values stored in the table fields changes according to
the tags of the table types.
We define the relation $\subtype_{c}$ as follows:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-FIELD1}\\
\dfrac{\senv \vdash V_{1} \subtype V_{2} \;\;\;
       \senv \vdash V_{2} \subtype V_{1}}
      {\senv \vdash V_{1} \subtype_{c} V_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FIELD2}\\
\dfrac{\senv \vdash V_{1} \subtype V_{2}}
      {\senv \vdash \Const \; V_{1} \subtype_{c} \Const \; V_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FIELD3}\\
\dfrac{\senv \vdash V_{1} \subtype V_{2}}
      {\senv \vdash V_{1} \subtype_{c} \Const \; V_{2}}
\end{array}
\end{array}
\]

These rules allow depth subtyping on $\Const$ fields.
The rule \textsc{S-FIELD1} defines that mutable fields are invariant,
while the rule \textsc{S-FIELD2} defines that immutable fields are covariant.
The rule \textsc{S-FIELD3} defines that it is safe to promote fields
from mutable to immutable.
We do not include a rule that allows promoting fields from immutable
to mutable because this would be unsafe due to variance.

There is a limitation on \emph{closed} table types that led us to
introduce \emph{open} and \emph{unique} table types.
If the table constructor had a \emph{closed} table type, then
programmers would not be able to use it to initialize a variable with
a table type that describes a more general type.
For instance,
\begin{verbatim}
    local t:{ "x":integer,
              "y":integer? } = { x = 1, y = 2 }
\end{verbatim}
would not type check, as the type of the table constructor would not
be a subtype of the type in the annotation.
More precisely,
\begin{align*}
& \{``x":1, ``y":2\}_{closed} \not\subtype \\
& \{``x":\Integer, ``y":\Integer \cup \Nil\}_{closed}
\end{align*}

Simply promoting the type of each table value to its supertype would
not overcome this limitation, as it still would give to the table constructor
a \emph{closed} table type without covariant mutable fields.
Thus, programmers would not be able to use the table constructor to
initialize a variable with a table type that includes an optional field.
Using the previous example,
\begin{align*}
& \{``x":\Integer, ``y":\Integer\}_{closed} \not\subtype \\
& \{``x":\Integer, ``y":\Integer \cup \Nil\}_{closed}
\end{align*}

We introduced \emph{unique} table types to avoid this limitation,
as they represent the type of tables with no keys that do not
inhabit one of the table's key types, and with no alias.
In particular, this is the case of the table constructor.
The following subtyping rule defines the subtyping relation among
\emph{unique} table types and \emph{closed} table types:
\[
\begin{array}{c}
\mylabel{S-TABLE2}\\
\dfrac{\begin{array}{c}
       \forall i \in 1..m \; \forall j \in 1..n \;
       \senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash V_{i} \subtype_{u} V_{j}' \\
       \forall j \in 1..n \; \not\exists i \in 1..m \;
       \senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
       \end{array}}
      {\senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{unique} \subtype
                    \{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{closed}}
\end{array}
\]

The rule \textsc{S-TABLE2} allows width subtyping and covariant keys.
It allows covariant keys because we also want to use \emph{unique}
table types as a way to join table fields that inhabit \emph{closed} table types.
For instance, we want to use the table constructor to initialize
a variable with a table type that describes a hash.

The rule \textsc{S-TABLE2} introduced the auxiliary relations
$\subtype_{u}$ and $\subtype_{o}$.
The first allows depth subtyping on all fields,
while the second allows the omission of optional fields.
We define them as follows:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-FIELD4}\\
\dfrac{\senv \vdash V_{1} \subtype V_{2}}
      {\senv \vdash V_{1} \subtype_{u} V_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FIELD5}\\
\dfrac{\senv \vdash V_{1} \subtype V_{2}}
      {\senv \vdash \Const \; V_{1} \subtype_{u} \Const \; V_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FIELD6}\\
\dfrac{\senv \vdash V_{1} \subtype V_{2}}
      {\senv \vdash V_{1} \subtype_{u} \Const \; V_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FIELD7}\\
\dfrac{\senv \vdash \Nil \subtype V}
      {\senv \vdash \Nil \subtype_{o} V}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FIELD8}\\
\dfrac{\senv \vdash \Nil \subtype V}
      {\senv \vdash \Nil \subtype_{o} \Const \; V}
\end{array}
\end{array}
\]

Using \emph{unique} table types to represent the type of the table
constructor allows our type system to type check the previous example.
More precisely,
\begin{align*}
& \{``x":1, ``y":2\}_{unique} \subtype \\
& \{``x":\Integer, ``y":\Integer \cup \Nil\}_{closed}
\end{align*}

Even though we allow width subtyping between \emph{unique} and \emph{closed}
table types, we do not allow it among \emph{unique} and other table types
because it would violate our definition of these other table types:
\[
\begin{array}{c}
\mylabel{S-TABLE3}\\
\dfrac{\begin{array}{c}
       \forall i \in 1..m \\
       \exists j \in 1..n \;
       \senv \vdash K_{i} \subtype K_{j}' \land \senv \vdash V_{i} \subtype_{u} V_{j}' \\
       \forall j \in 1..n \; \not\exists i \in 1..m \;
       \senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
       \end{array}}
      {\begin{array}{c}
       \senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{unique} \subtype\\
                    \{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{unique|open|fixed}
       \end{array}}
\end{array}
\]

The rule that handles subtyping between \emph{open} and \emph{closed} table
types allows width subtyping:
\[
\begin{array}{c}
\mylabel{S-TABLE4}\\
\dfrac{\begin{array}{c}
       \forall i \in 1..m \; \forall j \in 1..n \;
       \senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash V_{i} \subtype_{c} V_{j}' \\
       \forall j \in 1..n \; \not\exists i \in 1..m \;
       \senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
       \end{array}}
      {\senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{open} \subtype
                    \{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{closed}}
\end{array}
\]

>>>>

The rest of the paper is organized as follows:
Section \ref{sec:statistics} presents the methodology and results of
our survey of Lua idioms;
Section \ref{sec:types} presents the types of Typed Lua's
type system; Section~\ref{sec:subtyping} presents
their subtyping and consistent-subtyping relations;
Section \ref{sec:rules} presents the typing rules for evolution of table types,
and the typing rules for destructuring unions of tuples into projection types,
allowing flow typing on these projections;
Section \ref{sec:related} reviews related work;
Section \ref{sec:conclusion} presents our conclusions,
and outlines our plans for future work.

@@

\section{Types}
\label{sec:types}

@@

Second-level types are either tuple types or unions of tuple types, and represent sequences of values that appear as arguments, return values, and the left and right sides of
assignments. Tuple types are heterogeneous lists of first-class
types that end either with $\Void$ or with a {\em variadic type}.
A variadic type $F{*}$ represents an unbounded sequence of values of type $F \cup \Nil$.

@@

\section{Subtyping and Consistent-Subtyping}
\label{sec:subtyping}

Typed Lua's type system uses subtyping to order
types and  consistent-subtyping~\cite{siek2007objects,siek2013mutable}
to model compatibility between types in a way that allows
the interaction between statically and dynamically typed
parts of the program without collapsing the subtyping
relation.

The consistent-subtyping relation is defined in terms
of subtyping, so we focus the discussion in this section
on characterising the subtyping relation.

We present the subtyping rules as a deduction system for the
subtyping relation $\senv \vdash t_{1} \subtype t_{2}$.
The variable $\senv$ is a set of subtyping assumptions used
by the subtyping relation between recursive types.

Most of the subtyping rules are standard; we will focus
on the interesting cases, involving second-level types and
table types.

Tuple types are covariant:
\[
\begin{array}{c}
\mylabel{S-PAIR}\\
\dfrac{\senv \vdash F_{1} \subtype F_{2} \;\;\;
       \senv \vdash P_{1} \subtype P_{2}}
      {\senv \vdash F_{1} \times P_{1} \subtype F_{2} \times P_{2}}
\end{array}
\]

The semantics of Lua lets the right-hand side of an assignment
produce more values than l-values in the left-hand side,
with extra values silently dropped. It also lets the
right-hand side produce fewer values than l-values in the
left-hand side, using {\tt nil} to plug the holes. The
same behavior applies for lists of arguments vs. parameters.

Typed Lua preserves this behavior in assignment, but allows
both functions with strict arity (which use $\Void$) and
functions with flexible arity (which use variadic types).

The types of expression lists always end in a variadic type
that is $\Nil{*}$ if the number of values the list produces
is statically known. Rule \textsc{S-VARNIL-VOID} makes lets
an expression list with a statically known number of
values be used as the arguments or return values of a function
with a strict arity:
\[
\begin{array}{c}
\mylabel{S-VARNIL-VOID}\\
\senv \vdash \Nil{*} \subtype \Void
\end{array}
\]

The other rules let assignment and functions with flexible
arity reflect the semantics of Lua in a type-safe way. We
need rule {\sc S-VARIADIC} because just having variadic
types be covariant would be too restrictive (for example,
$\Nil{*}$ would not be a subtype of $\Number{*}$):
\[
\begin{array}{c}
\mylabel{S-VARIADIC}\\
\dfrac{\senv \vdash F_{1} \cup \Nil \subtype F_{2} \cup \Nil}
      {\senv \vdash F_{1}{*} \subtype F_{2}{*}}
\end{array}
\]

Rules {\sc S-VAR-EXPAND} and {\sc S-VAR-DROP} state the
conditions when variadic tuples if different length are
compatible:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-VAR-EXPAND}\\
\dfrac{\senv \vdash F_{1} \cup \Nil \subtype F_{2} \;\;\;
       \senv \vdash F_{1}{*} \subtype P_{2}}
      {\senv \vdash F_{1}{*} \subtype F_{2} \times P_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-VAR-DROP}\\
\dfrac{\senv \vdash F_{1} \subtype F_{2} \cup \Nil \;\;\;
       \senv \vdash P_{1} \subtype F_{2}{*}}
      {\senv \vdash F_{1} \times P_{1} \subtype F_{2}{*}}
\end{array}
\end{array}
\]

We could have the axioms $\senv \vdash F{*} \subtype F \cup \Nil \times F{*}$ and $\senv \vdash F \cup \Nil \times F{*}  \subtype F{*}$ instead of the two rules above, which can be derived
from these axioms by transitivity, but having the rules makes
the subtyping relation easier to implement.

We will now focus on the subtyping relation among table
types. Subtyping between a {\em fixed} or {\em closed}
table type and a {\em closed} table type resembles the
standard subtyping rule between mutable records, with
width subtyping plus depth subtyping from a mutable
or immutable field to an immutable field:
\[
\begin{array}{c}
\mylabel{S-TABLE-FC-C}\\
\dfrac{\begin{array}{c}
       \forall i \in 1..n \; \exists j \in 1..m \\
       \senv \vdash K_{j} \subtype K_{i}' \;\;\;
       \senv \vdash K_{i}' \subtype K_{j} \;\;\;
       \senv \vdash V_{j} \subtype_{c} V_{i}'
       \end{array}}
      {\begin{array}{c}
       \senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{fixed|closed} \subtype\\
                    \{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{closed}
       \end{array}} \; m \ge n
\end{array}
\]

Rule \textsc{S-TABLE-FC-C} uses the auxiliary relation $\subtype_{c}$ for handling depth subtyping:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-FIELD-C-MM}\\
\dfrac{\senv \vdash V_{1} \subtype V_{2} \;\;\;
       \senv \vdash V_{2} \subtype V_{1}}
      {\senv \vdash V_{1} \subtype_{c} V_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FIELD-C-II}\\
\dfrac{\senv \vdash V_{1} \subtype V_{2}}
      {\senv \vdash \Const \; V_{1} \subtype_{c} \Const \; V_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FIELD-C-MI}\\
\dfrac{\senv \vdash V_{1} \subtype V_{2}}
      {\senv \vdash V_{1} \subtype_{c} \Const \; V_{2}}
\end{array}
\end{array}
\]

The subtyping relation between {\em unique} and {\em closed}
table types is more relaxed, as the typing rules of Section~\ref{sec:tables} guarantee that a {\em unique}
table type can only appear on the left side of a subtyping
judgment if the reference to the table with the {\em unique}
tag is being destroyed:
\[
\begin{array}{c}
\mylabel{S-TABLE-U-C}\\
\dfrac{\begin{array}{c}
       \forall i \in 1..m \; \forall j \in 1..n \;
       \senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash V_{i} \subtype_{u} V_{j}' \\
       \forall j \in 1..n \; \not\exists i \in 1..m \;
       \senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
       \end{array}}
      {\senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{unique} \subtype
                    \{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{closed}}
\end{array}
\]

Rule {\sc S-TABLE-U-C} allows with subtyping, depth subtyping
on any field, not only immutable fields, as well as the addition
of new fields and the merging of existing fields, as long as the types of the values of new and merged fields include $\Nil$.
This rule uses the auxiliary relations $\subtype_{u}$ and $\subtype_{o}$, which respectively allow depth subtyping on all fields, and the addition of new fields:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-FIELD-U-MM}\\
\dfrac{\senv \vdash V_{1} \subtype V_{2}}
      {\senv \vdash V_{1} \subtype_{u} V_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FIELD-U-IM}\\
\dfrac{\senv \vdash V_{1} \subtype V_{2}}
{\senv \vdash \Const \; V_{1} \subtype_{u} V_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FIELD-U-II}\\
\dfrac{\senv \vdash V_{1} \subtype V_{2}}
      {\senv \vdash \Const \; V_{1} \subtype_{u} \Const \; V_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FIELD-U-MI}\\
\dfrac{\senv \vdash V_{1} \subtype V_{2}}
      {\senv \vdash V_{1} \subtype_{u} \Const \; V_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FIELD-O-M}\\
\dfrac{\senv \vdash \Nil \subtype V}
      {\senv \vdash \Nil \subtype_{o} V}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FIELD-O-I}\\
\dfrac{\senv \vdash \Nil \subtype V}
      {\senv \vdash \Nil \subtype_{o} \Const \; V}
\end{array}
\end{array}
\]

The rationale for rule {\sc S-TABLE-U-C} derives from the
fact that reading a non-existing key just yield $\Nil$
instead of being an error in Lua semantics, plus the fact
that the type system guarantees that a table with a
{\em unique} type has no keys with a value that does not
belong to one of its key types, and does not have any
aliases. A table with a type $\{1{:}\String, 2{:}\String,
``x"{:}\Integer, ``y"{:}\Value\}_{unique}$ can be safely recast as a
table with type 
$\{\Number{:}\String \cup \Nil, ``x"{:}\Number, ``z"{:}
\Number \cup \Nil\}_{closed}$ as long as it cannot be
accessed anymore with the original type.

Subtyping from an {\em open} table type to a {\em closed}
table type also allows width subtyping, merging fields,
and adding new fields, but forbids depth subtyping between
mutable fields or an immutable and a mutable field, as
the typing rules guarantee that the {\em open} reference
is destroyed, but the {\em open} table may already have {\em closed} aliases:
\[
\begin{array}{c}
\mylabel{S-TABLE-O-C}\\
\dfrac{\begin{array}{c}
	\forall i \in 1..m \; \forall j \in 1..n \;
	\senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash V_{i} \subtype_{c} V_{j}' \\
	\forall j \in 1..n \; \not\exists i \in 1..m \;
	\senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
	\end{array}}
{\senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{open} \subtype
	\{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{closed}}
\end{array}
\]

The subtyping relation from {\em unique} to {\em unique},
{\em open} or {\em fixed} table types is still flexible, but forbids standard width subtyping, as all these table types keep the guarantee that the values of all keys in tables of these types belong to one of their key types. Merging fields and adding new fields are still allowed:
\[
\begin{array}{c}
\mylabel{S-TABLE-U-UOF}\\
\dfrac{\begin{array}{c}
       \forall i \in 1..m \\
       \exists j \in 1..n \;
       \senv \vdash K_{i} \subtype K_{j}' \land \senv \vdash V_{i} \subtype_{u} V_{j}' \\
       \forall j \in 1..n \; \not\exists i \in 1..m \;
       \senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
       \end{array}}
      {\begin{array}{c}
       \senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{unique} \subtype\\
                    \{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{unique|open|fixed}
       \end{array}}
\end{array}
\]

The subtyping relation between {\em open} and {\em open} or
{\em fixed} table types mixes the depth subtyping restrictions of rule {\sc S-TABLE-O-C} with the width subtyping restrictions of
rule {\sc S-TABLE-U-UOF}, but still allow merging fields and
adding new fields:
\[
\begin{array}{c}
\mylabel{S-TABLE-O-OF}\\
\dfrac{\begin{array}{c}
       \forall i \in 1..m \\
       \exists j \in 1..n \;
       \senv \vdash K_{i} \subtype K_{j}' \land \senv \vdash V_{i} \subtype_{c} V_{j}' \\
       \forall j \in 1..n \; \not\exists i \in 1..m \;
       \senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
       \end{array}}
      {\begin{array}{c}
       \senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{open} \subtype\\
                    \{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{open|fixed}
       \end{array}}
\end{array}
\]

Finally, the subtyping relation between {\em fixed} table
types is the more restrictive, and does not allow width
subtyping, merging fields, or adding fields, just standard
depth subtyping:
\[
\begin{array}{c}
\mylabel{S-TABLE-F-F}\\
\dfrac{\begin{array}{c}
       \forall i \in 1..n \; \exists j \in 1..n \\
       \senv \vdash K_{j} \subtype K_{i}' \;\;\;
       \senv \vdash K_{i}' \subtype K_{j} \;\;\;
       \senv \vdash V_{j} \subtype_{c} V_{i}'
       \end{array}}
      {\senv \vdash \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{fixed} \subtype
                    \{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{fixed}}
\end{array}
\]

We will revisit the subtyping rules among table types on
Section~\ref{sec:tables}.

Subtyping between two recursive types uses the \emph{Amber rule}~\cite{cardelli1986amber}:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-AMBER}\\
\dfrac{\senv[x_{1} \subtype x_{2}] \vdash T_{1} \subtype T_{2}}
      {\senv \vdash \mu x_{1}.T_{1} \subtype \mu x_{2}.T_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-ASSUMPTION}\\
\dfrac{x_{1} \subtype x_{2} \in \senv}
      {\senv \vdash x_{1} \subtype x_{2}}
\end{array}
\end{array}
\]

Subtyping between a recursive type and a non-recursive type, or
vice-versa, uses a pair of rules that unfold the recursive type:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-UNFOLD-R}\\
\dfrac{\senv \vdash T_{1} \subtype [x \mapsto \mu x.T_{2}]T_{2}}
      {\senv \vdash T_{1} \subtype \mu x.T_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-UNFOLD-L}\\
\dfrac{\senv \vdash [x \mapsto \mu x.T_{1}]T_{1} \subtype T_{2}}
      {\senv \vdash \mu x.T_{1} \subtype T_{2}}
\end{array}
\end{array}
\]

The dynamic type $\Any$ is only a subtype of itself and of
$\Value$, and no type except $\Any$ is a subtype of $\Any$.
The type $\Any$ interacts with other types through the
{\em consistency} and {\em consistent-subtyping} 
relations~\cite{siek2007objects}. We use the symbol $\sim$ for
the consistency, and $\lesssim$ for consistent-subtyping.

The type $\Any$ is trivially consistent with all types, and all types
are consistent with $\Any$. A type is also consistent with
itself. Two union types $F_1 \cup F_2$ and $F_3 \cup F_4$ are
consistent if either $F_1 \sim F_3$ and $F_2 \sim F_4$ or
$F_1 \sim F_4$ and $F_2 \sim F_3$. Two function types
$S_1 \rightarrow S_2$ and $S_3 \rightarrow S_4$ are consistent
if and only if $S_1 \rightarrow S_3$ and $S_2 \rightarrow S_4$. Two table types are consistent if and only if they have the same key types, and the value types of each key type are consistent. Two recursive types are consistent if and only if their contents are consistent assuming that their type variables are consistent. Unions of tuple types follow the same rules as unions of first-class types. Variadic types are consistent if and only if their base types are consistent. 
Finally, pair types $F_1 \times P_1$ and $F_2 \times P_2$ are
consistent if and only if $F_1 \sim F_2$ and $P_1 \sim P_2$.

In a language with subtyping, the consistency relation is
too restrictive. We need to combine it with subtyping into a
{\em consistent-subtyping} relation. The definition is
straightforward: a type is a consistent-subtype
of another if the first type is a subtype of a type
that is consistent with the other, or if the first type
is consistent with a type that is a subtype of the other.

In the implementation of Typed Lua, we also use consistent-subtyping to normalize and simplify union types.
For instance, the union type $\Boolean \cup \Any$ simplifies
to $\Any$, because $\Boolean$ is a consistent-subtype
of $\Any$. Another example is the union type
$\Number \cup \Nil \cup 1$, that simplifies to
$\Number \cup \Nil$, because $1$ is a consistent-subtype of $\Number$.
